\addcontentsline{toc}{chapter}{Foreword}
\chapter*{Foreword}

Welcome to the rules for the \textit{Dreamshard RPG}!

My goals in this project were to create a system that had simple rules, but a lot of gameplay depth. I wanted to let the players break free of traditional, class-based system, and instead give a lot of flexibility to both the GM and the players. As such, Characters instead draw from Ability trees than can be mixed and matched. As such, the system also permits a lot of Spell customization, and the design of custom Gadgets. 

I also wanted to diminish the "dump stat" effect, by making every Characteristing potentially relevant. The use of a roll system with degrees of success (and the presence of partial successes called "Grazes") also significantly reduces random misses, resulting in lower variance of outcomes. During playtests, I found this resulted in a more consistent gameplay experience, and crucially makes gameplay a lot faster.

The current version, however, has its drawbacks. The system relies a lot on adjudication by the GM, in particular for the design and balancing of the Tools, Abilities and Spells to give the players. I hope the guidelines provided in the Balancing section can help alleviate this.

Finally, this entire project was an excuse to do some worldbuilding, which as any Game Master ultra-nerd will tell you, is always one of the best parts. I hope you will enjoy the results, and above all, \textbf{have fun!}



\addcontentsline{toc}{section}{Acknowledgements}
\section*{Acknowledgements}

I wish to thank the members of my gaming group for helping me beta-test this system. In alphabetical order: Jeanne Chèneby, Laurent Hannouche, Adrien Long, Justine Long an Florian Rosier. In particular, special thanks to Jeanne for her copious feedback on the manuscript and the design of the rules themselves.

Also thanks to the designers of the Pillars of Eternity paper RPG, which was a huge inspiration.

The LaTeX template used in this manuscrit is a derivation the one designed by \textit{evanbergeron}. The font used, Souvenir, is used as a non-commercial application.

\vskip 0.5cm

This work is made under an open \textit{Creative Commons-BY} license. 

\textit{This version of the manuscript was compiled on October 31st, 2023.}

Contact: \textit{quentin.q.ferre@gmail.com}



\paragraph{Art Credits}

In no particular order: Arnold Böcklin, Patrick O'Brien, Dariusz Zawadzki, Bob Ross, Mosè Bianchi, Dominique Van den Broech, Roland Davies, David and Nansi Gallup, Poulsen, Wonderdraft.


%% By file name :
% naval = Battle of the Chesapeake (Patrick O'Brien)
% machine head = Dariusz Zawadzki
% Island = bob ross
% death whisper = Self-Portrait with Death Playing the Fiddle (Arnold Böcklin)
% dark nun = Monaca di Monza, by Mosè Bianchi
% fort = Dominique Van den Broech
% galleon = roland davies
% kelp = David and Nansi Gallup
% port first choice = Poulsen